WORKING TITLE: The adaptive value of green nest material across birds: A systematic review and meta-analysis

### 2025-06-18 manuscript_GNM.qmd
Revising Introduction.. not rendered, just qmd file from one machine to another

Another hypothesis to explain the adaptive benefit of GNM is the widely tested nest protection hypothesis (NPH), which proposes that volatile compounds in green material reduce parasite abundance or pathogen loads, indirectly enhancing offspring survival. In an influential early comparitive study in raptors (Falconiformes), @wimberger1984 found indirect evidence for the reduction of ectoparasites in the nests with green nest material. @clark1988 later provided experimental support in European Starlings, showing that preferred herbs contained insect-repellent and antimicrobial volatile. They also outlined key criteria for evaluation of the NPH, including selective plant use, bioactive chemical properties and demonstrable adaptive benefits. Field observation and experimental studies on raptors [@ontiveros2007; @dykstra2009] and passerines [@dawson2004; @lafuma2001; @mennerat2009b; @quiroga2012] have shown mixed results, and in some cases, GNM improved nestling condition without reducing parasite abundance [@gwinner2000]. These findings led to the development of drug hypothesis, which suggests that the aromatic GNM directly enhance nestling development or immune function, independent of the parasite control [@gwinner2000; @mennerat2009a]. Thus, while both hypotheses posit adaptive benefits of GNM, the drug hypothesis emphasizes a pharmacological mechanism, and the two are not mutually exclusive. Several other adaptive functions of GNM have also been proposed. These functions do not refer to the aromatic properties of GM, but could still provide some benefits such as enhance the crypticity of the nest [@skutch1957], provide insulation for eggs and nestling [@mertens1977], provide protection from sun, signal nest occupancy or simply fill the gaps between nesting materials [@lyons1987]. These alternative functional roles have received little support or attention.

Despite several observational and experimental studies and two crucial narrative reviews [@dubiec2013; @scott-baumann2015], a comprehensive quantitative synthesis across studies and hypotheses has been lacking. This has limited our understanding of the generality and magnitude of GNM effects on fitness benefits. To address this gap, we conducted a systematic review and meta-analysis to assess the overall effect of GNM on various fitness proxies. We explicitly evaluate the evidence for the courtship and parental-care- the latter combining the nest protection and drug hypotheses due to their non-exclusivity. The drug hypothesis can only be supported by improved nestling health while ensuring that GNM has no protective effect against any parasite or pathogen.

In addition to testing these primary hypotheses, we further explore the context-dependency in the research field and examine the potential sources of variation in the results (heterogeneity) in the GNM effects. Specifically, we explored whether the effect of GNM varied depending on the type of parasite (arthropods vs. microorganisms), the timing of GNM addition (e.g., before hatching, after hatching, or continuously), the experimental design (e.g., aromatic vs. non-aromatic vs. no added material), the focal species, and the type of fitness trait measured (e.g., physiology, morphology, reproduction). These exploratory moderators could allow us to assess context dependence and identify ecological or methodological patterns that may explain inconsistencies in previous findings. Together, our evidence synthesis approach provides the first quantitative test of the adaptive value of green nest material across bird species and research contexts.

>>>>

Second and the most widely tested one, is the nest protection hypothesis, which proposes that the volatile compounds in GNM reduce the amount of parasites or pathogens in the nest, which could enhance offspring survival. This hypothesis traces back to a comparative study in raptors (Falconiformes) that proposed it and indirectly tested it by comparing the nest reuse, diet composition and the green nest material presence \[@wimberger1984\]. Soon after, a preference for GNM containing insect-repellent and antimicrobial volatile was experimentally confirmed in European Starlings, therefore confirming two of the key criteria for evaluating the nest protection hypothesis (i.e., selective plant use, bioactive chemical properties and demonstrable adaptive benefits; @clark1988). Observational and experimental studies on raptors \[@ontiveros2007; @dykstra2009\] and passerines [@dawson2004; @lafuma2001; @mennerat2009b; @quiroga2012] have shown mixed results, and in some cases, GNM has been shown to improve offspring condition without reducing the amount of parasites in the nests [@gwinner2000], which led to the development of drug hypothesis. The drug hypothesis suggests that the volatiles present in GNM directly enhance offspring development or immune function, independent of the parasite control [@gwinner2000; @mennerat2009a]. Thus, while both hypotheses suggest adaptive benefit of GNM by enhancing offspring survival, the drug hypothesis emphasizes a pharmacological mechanism, whereas the nest protection hypothesis emphasizes parasite and pathogen repellency. Importantly, these two hypothese are not mutually exclusive unless the improve in offspring health can be proved to be unrelated to a reduction in parasite or pathogen load.

Last, several additional adaptive hypotheses or functions of GNM that do not refer to its aromatic properties have been proposed. These include benefits such as enhancing the crypticity of the nest [@skutch1957], providing insulation or sun protection for eggs and offspring [@mertens1977], signal nest occupancy or simply fulfilling a structural function [@lyons1987]. However, these alternative functional roles have received little support or attention. In all, despite decades of research, abundant observational and experimental studies, and two crucial narrative reviews [@dubiec2013; @scott-baumann2015], we still do not know the overall adaptive significance of GNM, its generality across bird species, and the relative importance of each of the suggested mechanisms. Given the importance of nests and nest building across the tree of life, and their increasing recognition as an expression of the extended phenotype, it is of paramount importance to qualitatively synthesise this field.

To address this gap, we conducted a pre-registered systematic review and meta-analysis of experiments testing the effect of GNM on various fitness proxies across birds. Our meta-analysis explicitly evaluates the overall evidence for the adaptive significance of GNM, as well as to assess the relative importance of the main suggested hypothesis: the courtship hypotheses and the parental care hypothesis i.e. combining both the nest protection and drug hypotheses due to their non-exclusivity. In addition, we explore the observed context-dependency (i.e., high heterogeneity among effect sizes) via meta-regressions. Specifically, we tested whether the effect of GNM varied depending on the type of parasite (arthropods vs. microorganisms), the timing of GNM addition (before hatching, after hatching, or continuously), the experimental design (aromatic vs. non-aromatic vs. no added material), the focal species, and the type of fitness proxy investigated (physiology, morphology, reproduction). Our meta-analysis of experimental studies corresponds to the first quantitative synthesis testing the causal relationship between GNM and fitness across birds.

These exploratory moderators could allow us to assess context dependence and identify ecological or methodological patterns that may explain inconsistencies in previous findings. Together, our evidence synthesis approach provides the first quantitative test of the adaptive value of green nest material across bird species and research contexts.
